\section*{Abstract}

Concept erasure is useful only when it removes a target attribute without destroying the other information encoded in the representation. This is hard in the nonlinear setting: information erased for one probe may remain recoverable by a fresh nonlinear probe. Direct gradient-based edits introduce a second failure mode: they may remove target signal by pushing representations \emph{off the manifold} of natural representations, damaging the many other concepts encoded along that manifold --- most of which are unlabeled and cannot be explicitly protected as controls. Our central argument is that nonlinear erasure becomes surgical precisely when each edit is constrained to stay on the natural-representation manifold: an on-manifold step removes target signal in directions the representation itself supports, so collateral damage to unknown, unlabeled control concepts is bounded by construction. We propose \TanGCE{}, a manifold-aware method for nonlinear concept erasure that operationalises this argument. \TanGCE{} first removes dominant low-order target signal with closed-form projections that eliminate the main mean and covariance differences between the target classes. It then targets the remaining nonlinear signal locally on the manifold. For each representation, nearest-neighbor differences estimate the local tangent space of the manifold; \TanGCE{} projects the concept-scorer gradient onto this tangent space, scales each coordinate by the local manifold extent, and caps the update with a per-sample trust region. Thus, the erasure step is not merely nonlinear, but constrained to directions and scales supported by the local manifold of natural representations. We evaluate \TanGCE{} on 13 language models, three NLP concepts, and 40 CelebA-CLIP attributes under two control regimes, using one fixed set of hyperparameters with no per-setting tuning. Compared at the same allowed drop in control-concept accuracy, \TanGCE{} removes more target signal than the strongest published baseline. On the NLP settings, it drives a fresh nonlinear probe to the majority-class floor in $26$--$33$ settings, $13$--$20$ more than the baseline, while causing smaller control degradation. The advantage is largest when target and control attributes are entangled: on CelebA-CLIP, under highly correlated controls, \TanGCE{} is the only method that stays within the allowed control-accuracy drop for every target attribute --- the empirical signature of an erasure step that respects the manifold the controls also live on.
