% Appendix snapshot tables
\section{Single-Model Layer-8 Comparison (All Methods)}
\label{sec:appendix_full_comparison}

Table~\ref{tab:full_comparison} reports the full eight-method comparison at layer~8 of Qwen2.5-0.5B across sycophancy, gender, and safety. This table provides the detailed single-model benchmark that supports the multi-model summary in the main text.

\input{tables/full_comparison}

\section{CelebA three-concept comparison (legacy protocol)}
\label{sec:appendix_celeba_clip_fair}

Table~\ref{tab:celeba_clip_fair} reports the focused three-concept CelebA comparison (Male, Smiling, Blond\_Hair) on 5 least-correlation controls each, under the legacy global-$\lambda$ HPO protocol used by Tables~\ref{tab:celeba_clip_all40} and~\ref{tab:multi_model}. We retain it as a supplementary slice; the headline CelebA evidence is the 40-attribute, two-control-regime comparison in Table~\ref{tab:celeba_correlated_controls}.

\input{tables/celeba_clip_fair}

\section{All Methods, All Models, All Datasets}
\label{sec:appendix_all_methods}

Tables~\ref{tab:appendix_all_methods_sycophancy}, \ref{tab:appendix_all_methods_gender}, and~\ref{tab:appendix_all_methods_safety} report nonlinear leakage (NL) and control-task change ($\Delta$Ctrl) for all eight erasure methods across thirteen models spanning five families (Qwen2.5, Gemma-2, Llama-3.2, Mistral, Gemma-3) and all three concept domains. Values are mean\,$\pm$\,std over the 25\%, 50\%, and 75\% depth layers. The lowest NL per model among non-random methods is marked in \textbf{bold}. Gemma-3 coverage now includes completed 1B, 4B, 12B, and 27B results in the aggregate tables; the per-layer table below is narrower and should be read as the supporting breakdown for the models currently mirrored there.

\input{tables/appendix_all_methods}

\section{Per-Layer Multi-Model Breakdown}
\label{sec:appendix_per_layer}

Table~\ref{tab:multi_model_per_layer} gives the full per-layer results underlying the mean\,$\pm$\,std values in Table~\ref{tab:multi_model}. Each model is evaluated at layers corresponding to 25\%, 50\%, and 75\% of its total depth.

\input{tables/multi_model_per_layer}

\section{Supplementary Deeper-Layer Comparisons}
\label{sec:appendix_snapshots}

This appendix records broader deeper-layer comparisons that are directionally consistent with the main text but are not part of the core quantitative evidence used for the paper's principal claims. We include them to show how the same comparison pattern extends beyond the main layer-8 evaluation and to motivate future matched-leakage analyses.

\subsection{Iterative Sycophancy (Supplementary Comparison)}
\input{tables/repo_snapshot_sycophancy}

\subsection{Iterative Safety (Supplementary Comparison)}
\input{tables/repo_snapshot_safety}

\subsection{Local CelebA Visual Table-2-Style Comparison}
\input{tables/local_celeba_visual_table2}

To provide a modality-matched supplementary slice, this appendix reports a local vision-only CelebA run (CLIP ViT-B/32 embeddings) over the same method family, with three least-correlated controls per target concept. This local check is separate from the paper-locked main-text protocol and is included only as supplementary context.

\subsection{CelebA Per-Concept Detail (All 40 Attributes)}
\label{sec:appendix_celeba_per_concept}
Tables~\ref{tab:celeba_clip_all40_per_concept_leak} and~\ref{tab:celeba_clip_all40_per_concept_ctrl} give the full per-concept breakdown behind the aggregate CelebA result in Table~\ref{tab:celeba_clip_all40}. The main-text Table~\ref{tab:celeba_clip_fair} reports three representative concepts (Male, Smiling, Blond\_Hair); the appendix tables cover all 40 binary CelebA attributes from the same CelebA all-40 evaluation setting under the shared validation-HPO protocol. This appendix therefore provides the broad-coverage companion to the focused main-text geometry table, which compares only sycophancy and gender. \TanGCE{} achieves the lowest nonlinear leakage on \textbf{36 of 40} concepts (LEACE wins on 3, IGBP on 1), with a mean margin of $0.116$ absolute NL accuracy over the best competing method per concept. Per-concept mean $\Delta$Ctrl remains within roughly a percentage point of zero, consistent with the aggregate mean NL of $0.707$ and mean $\Delta$Ctrl of $-0.7\%$ reported in the main text.
\input{tables/celeba_clip_all40_per_concept}

\input{tables/main_text_geometry_schedule}

\paragraph{Geometry Note.}
The broader geometry report continues to support three useful points: early and mid-layer representations are effectively low-dimensional, on-manifold directions carry much of the model's sensitivity in these perturbation tests, and interpolation remains consistent with a connected representation space. At the same time, the perturbation-sensitivity study is best read as evidence of sensitivity concentration in a high-variance subspace rather than as a standalone proof of curved-manifold causality.

\section{Intuition: ``fraction of the local whitened neighborhood''}
\label{sec:appendix_fmaha_intuition}

Section~\ref{sec:closed_form_lambda} introduces the per-sample trust region $\| \tilde{\mathbf{x}}_i - \mathbf{x}_i \|_2 \leq \varepsilon \cdot s_i$ and offers two choices for the local scale $s_i$: the simple radial $k$-NN distance $r_i$, or the anisotropy-aware whitened-tangent radius (Eq.~\ref{eq:s_fmaha}). The paper says $\varepsilon$ then becomes ``a fraction of the local whitened-tangent neighborhood.'' This appendix unpacks what that phrase means in concrete terms.

\paragraph{Reading the trust region as a budget.}
With $s_i = r_i$ (the radial choice), the bound reads ``move at most $\varepsilon \cdot$ Euclidean $k$-NN radius.'' This is direction-agnostic: every direction inside ambient space gets the same budget. With $s_i$ defined by Eq.~\ref{eq:s_fmaha} (the FMaHa whitened-tangent radius), the bound instead reads ``move at most $\varepsilon$ of the neighborhood radius \emph{after normalising each local tangent axis by its own scale}.'' That distinction is what makes $\varepsilon$ direction-aware.

\paragraph{Worked example.}
Suppose the local tangent space at $\mathbf{x}_i$ is two-dimensional with right singular vectors $\mathbf{v}_1, \mathbf{v}_2$ and singular values $\sigma_{i,1} = 10$, $\sigma_{i,2} = 1$. Neighbors of $\mathbf{x}_i$ extend ten times further along $\mathbf{v}_1$ than along $\mathbf{v}_2$:
\[
\text{$\mathbf{v}_1$ direction: long / fat / well-supported,} \qquad
\text{$\mathbf{v}_2$ direction: short / thin / fragile.}
\]
The whitened tangent coordinate of a neighbor offset $\mathbf{x}_j - \mathbf{x}_i$ along $\mathbf{v}_k$ is
$z_k = \langle \mathbf{x}_j - \mathbf{x}_i, \mathbf{v}_k \rangle / \sigma_{i,k}$. Moving 1 ambient unit along $\mathbf{v}_1$ is therefore $1/10 = 0.1$ \emph{whitened} units, while moving 1 ambient unit along $\mathbf{v}_2$ is $1/1 = 1$ whitened unit. Expressed in whitened coordinates the local neighborhood is approximately spherical: ``10\% of the whitened neighborhood'' is the same allowed move \emph{regardless of direction inside the tangent}. Translated back through the singular spectrum, that allowed move is large in ambient norm along $\mathbf{v}_1$ and small along $\mathbf{v}_2$ — exactly the asymmetry we want, because the linear approximation is reliable along high-$\sigma$ axes and fragile along low-$\sigma$ ones.

\paragraph{Comparison with the Euclidean radius.}
The Euclidean form is the inequality
\[
\| \boldsymbol{\delta} \|_2 \;\leq\; \varepsilon \cdot r_i,
\]
which treats every direction equally. The whitened form is approximately
\[
\Big\| \mathbf{B}_i^\top \boldsymbol{\delta} \,/\, \boldsymbol\sigma_i \Big\|_2 \;\leq\; \varepsilon \cdot R_i^{\mathrm{white}},
\]
where division is component-wise and $R_i^{\mathrm{white}}$ is the median neighbour displacement in standardised tangent coordinates (Eq.~\ref{eq:s_fmaha}). Reading this inequality in plain English: \emph{after dividing each tangent coordinate of the edit by its local standard scale $\sigma_{i,k}$, the edit must be at most $\varepsilon$ times the typical neighbour displacement on that scale}. The same edit, projected onto a high-$\sigma$ axis, is allowed a larger ambient stride; projected onto a low-$\sigma$ axis, it is held to a much smaller stride.

\paragraph{Mental model.}
Imagine the local neighborhood as an ellipse stretched along its high-$\sigma$ axis and pinched along its low-$\sigma$ axis. A ``$\varepsilon$ Euclidean'' step ignores the ellipse shape — it steps the same length in any direction. A ``$\varepsilon$ whitened-neighborhood'' step instead asks: \emph{after reshaping the ellipse into a sphere, does the edit fit inside an $\varepsilon$ ball?} FMaHa then maps the allowed move back into the original representation space. The result is a direction-aware trust region: it prevents the iterative loop from taking a step that is numerically small in Euclidean norm but huge relative to a thin local tangent direction where the linear approximation is the least reliable.

This is the geometric reason why $\varepsilon = 0.1$ is interpretable as an order-of-magnitude smaller knob than the legacy global $\lambda$ used in the appendix tables: it bounds the edit relative to the local neighborhood's natural scale on each axis, not to a global step size.

\paragraph{Why $\varepsilon$ transfers across datasets without per-panel tuning.}
A direct consequence of the whitening is that $\varepsilon$ no longer carries the units of the local representation geometry: each tangent axis is rescaled by its own $\sigma_{i,k}$ before $\varepsilon$ is applied, so a representation panel with a wide local spread along its dominant axes and one with a narrow local spread both report the same allowed move when expressed as a fraction of their respective whitened neighborhoods. This is why we can fix $\varepsilon = 0.1$ once and use it across all 39 NLP panels (13 language models $\times$ 3 concepts at 50\%-depth) and all 80 CelebA panels (40 attributes $\times$ 2 control sets) without any per-panel tuning of the trust-region radius. Crucially, this transferability requires that the per-axis scale $\sigma_{i,k}$ \emph{itself} be estimated from the local representation — exactly what the local SVD on the neighborhood-secant matrix provides. The legacy global $\lambda$ in §\ref{sec:evaluation} cannot have this property because its units are panel-specific: the same $\lambda$ moves a Gemma-3-27B representation by a much larger ambient distance than it does a Qwen2.5-0.5B one, which is why the legacy protocol must run a per-panel $\lambda$-HPO grid and ours does not.

\section{CelebA control-set construction (least- vs.\ most-correlated five attributes)}
\label{sec:appendix_celeba_controls}

For each of the 40 binary CelebA attributes used as a target concept, we report two complementary control regimes (Section~\ref{sec:celeba}, Table~\ref{tab:celeba_correlated_controls}, Figure~\ref{fig:per_attr_bars_celeba}): the \emph{5 least-correlated} attributes and the \emph{5 most-correlated} attributes, both selected from the remaining 39 attributes. This appendix records exactly how the two control sets are computed.

\paragraph{Computation.}
Let $\mathbf{Y} \in \{0, 1\}^{N_{\text{train}} \times 40}$ be the train-split label matrix, with $\mathbf{Y}_{i,k}$ the label of training example $i$ on attribute $k$. We compute the empirical Pearson correlation matrix
\begin{equation}
\mathbf{R}_{k\ell} \;=\; \frac{\mathrm{Cov}(\mathbf{Y}_{:,k}, \mathbf{Y}_{:,\ell})}{\sqrt{\mathrm{Var}(\mathbf{Y}_{:,k}) \cdot \mathrm{Var}(\mathbf{Y}_{:,\ell})}}
\quad \text{for } k, \ell \in \{1, \ldots, 40\}.
\end{equation}
For each target attribute $k$, we then rank the remaining $39$ attributes by absolute correlation $|\mathbf{R}_{k\ell}|$:
\begin{itemize}
\item the \emph{5 least-correlated} controls are the bottom 5 by $|\mathbf{R}_{k\ell}|$ — they share \emph{minimal} information with the target and isolate clean surgicality;
\item the \emph{5 most-correlated} controls are the top 5 by $|\mathbf{R}_{k\ell}|$ — they share \emph{substantial} information with the target and form a stricter test of whether erasure can preserve genuinely related downstream features.
\end{itemize}
The target attribute itself is excluded from the candidate pool (i.e., $\ell \neq k$). Pairs are computed in float64 with no shrinkage. The full $40 \times 40$ correlation matrix and the resulting per-target control sets are stored in \texttt{outputs/celeba\_correlated\_controls.json} (committed for reproducibility) by \texttt{scripts/celeba\_compute\_correlations.py}.

\paragraph{Concrete examples.}
For \emph{Male}, the 5 most-correlated attributes are \emph{Wearing\_Lipstick} ($|r| \approx 0.78$), \emph{No\_Beard}, \emph{Heavy\_Makeup}, \emph{Attractive}, and \emph{5\_o\_Clock\_Shadow}; the 5 least-correlated are \emph{Narrow\_Eyes}, \emph{Blurry}, \emph{Straight\_Hair}, \emph{Receding\_Hairline}, and \emph{Pale\_Skin} (all $|r| < 0.05$). Erasing \emph{Male} while preserving the most-correlated controls is genuinely hard because the controls are partial proxies for the target; erasing while preserving the least-correlated controls tests whether the intervention is geographically narrow but not whether it disentangles correlated information. We therefore report both regimes side by side: each one isolates a different failure mode.

\paragraph{Why two regimes per concept.}
Reporting only one control set hides whichever regime the method happens to be weak in. A method that destroys the entire representation will pass the least-correlated test (controls are essentially independent of the edit anyway) and fail the most-correlated test catastrophically; conversely, a method that performs no erasure will pass both. The difference between the two regimes per method is therefore informative even when both individual numbers look acceptable. Section~\ref{sec:celeba} reads the difference as a measure of how much an erasure method confuses ``avoid this concept'' with ``avoid everything correlated with it''.

\section{Anisotropy regimes for $\mathbf{C}_i^{\alpha}$}
\label{sec:appendix_alpha_regimes}

The $\sigma$-preconditioned update of Eq.~\ref{eq:tangent_erase_cov} reweights each tangent direction by $\sigma_{i,k}^{2\alpha}$, where $\alpha \geq 0$ is an anisotropy exponent. Three values of $\alpha$ have natural interpretations:
\begin{itemize}
\item \textbf{$\alpha = 0$}: $\mathbf{C}_i^{\,0} = \mathbf{I}$, the update reduces to Eq.~\ref{eq:tangent_erase}. Reported as an ablation.
\item \textbf{$\alpha = \tfrac{1}{2}$}: a \emph{mild} anisotropy. Each direction is reweighted by $\sigma_{i,k}$ rather than by $\sigma_{i,k}^2$, preserving information from low-variance directions while still respecting the dominant geometry. \textbf{This is the default in this paper.}
\item \textbf{$\alpha = 1$}: full Mahalanobis preconditioning. The update direction $\mathbf{C}_i \mathbf{v}_i$ is the first-order optimum of $\min_{\boldsymbol\delta} \langle \mathbf{g}_i, \boldsymbol\delta \rangle$ subject to the Mahalanobis trust region $\boldsymbol\delta^\top \mathbf{C}_i^{-1} \boldsymbol\delta \leq \varepsilon^2$. Reported as an ablation; risks suppressing low-variance signal that may carry residual concept information.
\end{itemize}

\paragraph{Relation to Mahalanobis whitening (and what $\alpha = \tfrac{1}{2}$ is \emph{not}).}
A reader familiar with Mahalanobis whitening might wonder whether $\alpha = \tfrac{1}{2}$ is ``the whitening'' we are doing. It is not. Mahalanobis whitening transforms the data so the covariance becomes identity, $\mathbf{x}^{\mathrm{white}} = \mathbf{C}^{-1/2} \mathbf{x}$, and corresponds to the \emph{negative} half-power of $\mathbf{C}$ (i.e.\ $\alpha = -\tfrac{1}{2}$ in our notation). Whitening makes every direction equal-weighted, which would actively dilute the concept-specific signal we want to suppress and force every tangent axis to receive equal step magnitude.

We use $\mathbf{C}^{+1/2}$ (positive half-power), which does the opposite: it reweights the tangent step \emph{toward} the high-$\sigma_{i,k}$ principal axes — exactly the directions where the local manifold has the most support and where the linear approximation is most reliable — and away from low-$\sigma$ thin tail directions where the linear approximation is fragile and where motion most easily leaves the manifold. The three values $\alpha \in \{0, \tfrac{1}{2}, 1\}$ thus form a one-parameter family interpolating between (i) treating every tangent direction as equally trustworthy ($\alpha = 0$, isotropic), (ii) respecting the local ellipsoid mildly with $\sigma_{i,k}$ scaling ($\alpha = \tfrac{1}{2}$, our default), and (iii) the natural-gradient / Mahalanobis trust-region optimum that maximally suppresses thin directions ($\alpha = 1$). Whitening ($\alpha = -\tfrac{1}{2}$) sits outside this useful range for our problem because it has the opposite sign of the geometric correction we want; we therefore do not include it.

