% Conclusion
\section{Conclusion}

We introduced TanGCE, a manifold-aware concept erasure method that constrains erasure gradients to the local tangent space of the representation manifold. Under a fair validation-based hyperparameter protocol---where all methods search the same $\lambda$ grid---TanGCE achieves the lowest nonlinear leakage across three concept domains in Qwen2.5-0.5B (sycophancy, gender, and safety) while preserving control-task accuracy and maintaining near-zero feature redistribution.

The $\lambda$ sweep reveals a critical failure mode of ambient erasure: at high erasure strength, concept mutual information \emph{increases} because off-manifold displacement creates new concept-correlated axes of variation. TanGCE avoids this by construction, yielding monotonically decreasing leakage with increasing $\lambda$. This explains a key asymmetry in the fair comparison: TanGCE consistently selects $\lambda^*{=}8.0$ (the maximum in our grid), while Ambient GCE is constrained to $\lambda^*{=}0.5$---not because tangent projection is given an unfair advantage, but because it can safely scale where ambient methods cannot.

Cross-modal experiments on 40 CelebA facial attributes in CLIP ViT-B/32 representations confirm that TanGCE generalizes beyond language models. TanGCE wins 37 of 40 attribute erasure tasks under fair HPO, reducing mean nonlinear leakage to 0.707 (vs.\ 0.839 for LEACE, the strongest baseline), with comparable control-task preservation. The three exceptions are near-chance concepts where all methods perform similarly.

Our SAE-based diagnostics reveal that erasure evaluation must account for polysemanticity: at $8\times$ dictionary expansion, marginal correlations miss concept features that are polysemantic, reporting 0\% removal even when nonlinear probes confirm substantial erasure. At $64\times$ expansion, features become monosemantic and the SAE metrics align with probe-based evaluation.

Several directions remain open. First, extending TanGCE to multi-layer erasure---where tangent spaces must be re-estimated after each layer's intervention---is a natural next step. Second, while our experiments use PCA-based tangent estimation, learned manifold representations (e.g., variational autoencoders) could provide more accurate tangent spaces. Finally, the connection between local geometric structure and erasure effectiveness suggests that manifold-aware methods may benefit other representation editing tasks beyond concept removal.
