% Discussion
\section{Discussion}

The current results support a strong central claim, though not the strongest imaginable one. In every finished comparison, the tangent-based method removes more measured concept signal than its ambient counterpart, and across the broader appendix tables it is also stronger than the earlier nonlinear and linear baselines we compare against. What the present evidence does not yet show is the still stronger claim that tangent projection already guarantees the best point on the full leakage-utility frontier at every operating point. In the completed runs, tangent reaches lower leakage by taking larger steps, and its utility metrics are either slightly worse or mixed in some settings.

The geometry diagnostics still matter because they explain why sycophancy is the harder case. Sycophancy shows larger angular separation near $80^\circ$, while gender remains near $53^\circ$, so ambient edits are more likely to miss the local support of the primary nonlinear concept. That is consistent with tangent-based methods improving removal in both concepts while leaving sycophancy as the clearly more nonlinear case. It also helps explain why the random baselines can sometimes lower leakage only by corrupting the representation much more broadly: they act in ambient space with no mechanism for staying on the support that natural representations actually occupy.

\paragraph{Connection to low-rank fine-tuning.}
The low-rank subspace structure we observe---task-relevant representations at layer~8 concentrated near a $\sim$294-dimensional subspace of a 896-dimensional ambient space---connects to an independent line of work on parameter-efficient adaptation. \citet{aghajanyan-etal-2021-intrinsic} show that language model fine-tuning objectives have an intrinsic dimensionality of order $200$--$1{,}000$: optimizing the full parameter space is statistically equivalent to optimizing a random projection of it down to $m \ll d$ dimensions. \citet{hu-etal-2022-lora} exploit this structure directly via LoRA, approximating weight updates as low-rank matrices whose rank captures the intrinsic dimensionality of the task signal. Our finding is the activation-space analogue of the same phenomenon: just as parameter-space fine-tuning updates have low effective rank, the hidden-state directions that carry task-specific information occupy a much smaller local support than the full ambient space. The two observations are mechanistically linked: if layer-$\ell$ activations have intrinsic rank $m$, the gradients of any task-specific loss with respect to the weight matrices at that layer are also rank-$m$ at the activation level, because they are formed as outer products of the low-rank activations and the corresponding error signals. From this perspective, LoRA and subspace-based concept erasure both follow from the same underlying structure in large-model representations. The key distinction is not global versus local directionality, but whether the update is allowed to act freely in ambient space or is constrained to the manifold support that the representations actually occupy.

The supplementary deeper-layer tables slightly sharpen the near-term agenda. In Tables~\ref{tab:repo_snapshot_sycophancy} and \ref{tab:repo_snapshot_safety}, the same removal advantage continues at layers 16 and 23 for both sycophancy and safety, although the utility metrics differ by domain. We do not elevate those rows to headline claims because they are supplementary to the main layer-8 evaluation, but they strengthen the case for spending the next round of effort on matched-leakage comparisons and deeper-layer replications rather than on searching for a different primary mechanism.

Two implications follow. First, the paper should presently avoid claiming that manifold-aware control is already proven to improve the full leakage-utility frontier; the matched-leakage analysis and broader replications are still needed for that claim. Second, the current evidence suggests that the next important question is not whether tangent projection changes the update at all, but under which layers and operating points that removal advantage can be converted into a cleaner utility tradeoff.

The matched layer-8 linear-family comparison clarifies one part of that tradeoff. LEACE is more surgical than tangent on this slice because it removes linearly decodable concept signal while taking a much smaller geometric step and preserving more downstream utility, especially on sycophancy where the answer-matching penalty is substantially smaller. That does not make LEACE a substitute for tangent on the main nonlinear claim: LEACE succeeds by solving the linear erasure problem cleanly, whereas tangent remains the stronger method when the target concept is evaluated with a fresh nonlinear probe. The current evidence therefore suggests a division of labor rather than a single winner: LEACE is the cleaner linear baseline, while tangent remains the more aggressive nonlinear remover.
