% Experimental setup details
\subsection{Datasets, Models, and Compared Methods}

This subsection completes the experimental setup by describing the datasets, models, and comparison methods used to evaluate \TanGCE{}.

\subsubsection{Experimental Domains}
We evaluate our methods across four evaluation settings:
\begin{enumerate}
    \item \textbf{Sycophancy (Primary Nonlinear Case):} Our central focus is on removing agreement bias in model responses. We use the Anthropic model-written-evals sycophancy suite \citep{perez-etal-2022-model-written}, which comprises three subsets: NLP survey, PhilPapers 2020, and political typology quiz. Each example is a multiple-choice question prefaced by a persona bio that asserts a position, with a designated \emph{sycophantic} answer (the one matching the persona's assertion) and one or more \emph{non-sycophantic} alternatives. Because the dataset does not provide pre-assigned per-example concept labels, we assign them by running the model itself: concept~$=1$ if the model's argmax prediction matches the sycophantic answer, concept~$=0$ otherwise. This labelling strategy identifies exactly which examples activate the sycophancy direction in the model being studied, rather than relying on a fixed human annotation. For the surgicality control we use \emph{answer matching}: whether the model selects the non-sycophantic (factually correct) answer. This adapts the standard helpfulness control from safety datasets to the sycophancy setting, where ``helpful'' behaviour is operationalized as resisting social pressure and answering correctly.

    \item \textbf{Safety (Secondary Case Study):} We evaluate the surgicality of erasure on content safety using the PKU-SafeRLHF dataset \citep{ji-etal-2024-pku-saferlhf}, testing the preservation of helpfulness alongside safety leakage.
    \item \textbf{Bias in Bios (Diagnostic Baseline):} We use the Bias in Bios dataset \citep{dearteaga-etal-2019-bias} for gender and profession classification as a replication anchor. This setting provides a linear comparison point to contrast against the nonlinear gains of tangent-constrained erasure.
    \item \textbf{CelebA Visual Attributes (Cross-Modal Generalization):} We apply the same erasure methods to CLIP ViT-B/32 \citep{radford-etal-2021-clip} pooled embeddings (768 dimensions) of CelebA \citep{liu-etal-2015-celeba} face images. CelebA provides 40 binary facial attributes, enabling a rich surgicality analysis: when erasing one attribute (e.g., \emph{Male}), the remaining 39 serve as potential controls. We select the 5 least-correlated attributes as controls per concept to ensure that control accuracy changes reflect genuine collateral damage rather than attribute correlation.

\end{enumerate}

\subsubsection{Models}
\paragraph{Language models.} The main seven-method comparison is run on Qwen2.5-0.5B \citep{qwen-2024-qwen25} (24 decoder layers, hidden dimension 896) at layer~8. To test whether the tangent-vs.-ambient gap persists across families and scales, we additionally compare \TanGCE{} and \AmbGCE{} on Qwen2.5-1.5B, Qwen2.5-3B, Gemma-2-2B, Gemma-2-9B, Gemma-2-27B, Llama-3.2-1B, Llama-3.2-3B, Mistral-7B-v0.1, Gemma-3-1B, Gemma-3-4B, Gemma-3-12B, and Gemma-3-27B. \textbf{For each model we extract residual-stream representations from the layer at $50\%$ of total decoder depth} (e.g., layer~12 of 24 for Qwen2.5-0.5B, layer~16 of 32 for Mistral-7B, layer~24 of 48 for Gemma-3-27B). We pick mid-depth because (i) early layers encode mostly token-level features unrelated to the high-level concepts we erase, (ii) late layers risk degenerate task-specific representations that interact unpredictably with the LM head when re-injected, and (iii) on a held-out validation panel the $50\%$ layer was best or tied for both control utility and concept recoverability across all families we tested. The full Pareto comparison against Obliviator (Section~\ref{sec:pareto_obliviator}, Figure~\ref{fig:pareto_sota_obliv_floor}) reports the 39-panel grid (13 LLM families $\times$ 3 concepts) at this $50\%$-depth layer.

\paragraph{Vision encoder.} The CelebA and FairFace experiments use CLIP ViT-B/32 pooled embeddings (768 dimensions), extracted once and cached. We keep the encoder fixed across the two vision benchmarks so that the comparison isolates dataset effects rather than encoder differences.

\subsubsection{Compared Methods}
\paragraph{Methods.} The core quantitative comparison evaluates \TanGCE{} against eight baselines and ablations, plus a Majority Class reference floor: two random baselines (global and per-sample direction), linear projection, INLP \citep{ravfogel-etal-2020-null}, LEACE \citep{belrose-etal-2023-leace}, IGBP \citep{iskander-etal-2023-shielded}, Obliviator \citep{akbari-etal-2026-obliviator}, and \AmbGCE{}. \textbf{Obliviator} is the current state-of-the-art unsupervised concept-erasure baseline: an HSIC-encoder + RFF-EVP iterative projection that removes the empirically estimated nonlinear concept signal without using control labels. IGBP is the closest prior \emph{supervised} nonlinear baseline and serves as the main methodological reference point for the GCE family: it iteratively trains two-class MLP probes and applies Newton-step boundary corrections. We also report the majority-class predictor as a trivial recoverability floor for interpreting probe accuracy, although it is not itself an erasure method. \AmbGCE{} is our direct ablation of \TanGCE{}: it uses the same scorer-training loop but applies the update in the full ambient space. \TanGCE{} adds the key ingredient of the paper by estimating local tangent spaces with PCA-based neighborhoods and restricting the erasure direction to a local on-manifold approximation rather than the full ambient space.

\paragraph{Comparison method definitions.}
All erasure methods share a common goal: given representations $\mathbf{x}_i$ that encode concept $\phi$, produce modified representations $\tilde{\mathbf{x}}_i$ from which $\phi$ cannot be recovered, while preserving information unrelated to $\phi$.

\paragraph{Linear projection.}
Linear projection removes a single concept direction, typically estimated from the class-conditional means. It is the simplest global linear baseline and cannot remove nonlinear residual structure.

\paragraph{INLP (Iterative Nullspace Projection).}
INLP \citep{ravfogel-etal-2020-null} extends linear projection by repeatedly training linear classifiers on the residual and removing newly exposed linear directions. It is the canonical iterative linear baseline.

\paragraph{LEACE (Least-squares Concept Erasure).}
LEACE \citep{belrose-etal-2023-leace} is the strongest closed-form linear baseline in our comparisons. It performs a minimal linear intervention while eliminating linearly decodable concept information.

\paragraph{IGBP (Iterative Gradient-Based Projection).}
IGBP \citep{iskander-etal-2023-shielded} is the closest prior nonlinear baseline. It iteratively trains a nonlinear probe and moves each representation toward the current decision boundary by a Newton-style correction.

\paragraph{Obliviator.}
Obliviator \citep{akbari-etal-2026-obliviator} is the current state-of-the-art unsupervised nonlinear concept-erasure method. It alternates between training an HSIC-loss encoder that compresses the concept signal into a low-dimensional code and applying a random-Fourier-features EVP projection that removes that code from the representation. We use the published paper hyper-parameters: a 256-dim SiLU encoder trained with AdamW (lr~$=5{\times}10^{-4}$, weight decay $10^{-3}$) for 30 epochs on the first round and 25 epochs thereafter, run for $n=30$ iterations.

\paragraph{\AmbGCE{}.}
\AmbGCE{} is the direct ambient-space ablation of \TanGCE{}. It keeps the same nonlinear scorer loop but applies the update in the full hidden space rather than projecting it onto a local tangent approximation.

\input{tables/method_comparison_scope}

\subsubsection{Hyperparameter selection for the proposed pipeline}
\label{sec:hp_selection}

A consequence of the per-sample trust-region formulation in §\ref{sec:closed_form_lambda} is that the proposed \TanGCE{} pipeline has \emph{no $\lambda$ hyperparameter to sweep}. The step magnitude on each sample is determined in closed form from the trust-region radius $\varepsilon$ and the per-sample local geometry by Eq.~\ref{eq:lambda_closed_form}. We therefore do not run a validation HPO grid over $\lambda$ for any of the headline results in §\ref{sec:pareto_obliviator} (the 39-panel NLP comparison) or §\ref{sec:celeba} (the CelebA dual-controls comparison). Instead, we fix the trust-region radius once at $\varepsilon = 0.1$ and let $\lambda_i$ adapt per row.

\paragraph{What is and is not selected.}
For the headline pipeline we fix all step-controlling hyperparameters across panels: $\varepsilon = 0.1$, scorer hidden size $h = 512$, scorer training steps $= 800$, scorer-refit period $= 8$ outer iterations, $T = 60$ outer iterations, intrinsic-dimension floor $k_{\min} = 8$ (with $k_i$ from TwoNN per panel), and safety cap $\lambda_{\max} = 64$. The Phase-1 closed-form preprocessing (LEACE eraser + rank-2 $\Delta\Sigma$ projection, §\ref{sec:moment_preproc}) has no tunable hyperparameter beyond the projection rank, which we fix at $K = 2$ across all panels. We do not sweep any of these on a per-panel validation set: every panel — every (model, concept) cell in §\ref{sec:pareto_obliviator} and every (concept, control-set) cell in §\ref{sec:celeba} — uses the same hyperparameter values.

\paragraph{Where a $\lambda$ grid is still used.}
The $\lambda$ grid that appears in §\ref{sec:multi_model} (the 13-model \AmbGCE{}-vs-\TanGCE{} comparison) and in Table~\ref{tab:celeba_clip_all40} (the 40-attribute method comparison) is the legacy \emph{global-$\lambda$} form of the iterative loop, retained only for the ablation against \AmbGCE{} and for parity with prior published methods that share a single global step size. There a global $\lambda^* \in \{0.25, 0.5, 1.0, 2.0, 4.0, 8.0, 16.0, 32.0\}$ is selected per panel by lowest validation nonlinear-probe accuracy; this gives those baselines an HPO advantage, since the proposed pipeline reports results at a fixed $\varepsilon$. Both the canonical global-$\lambda$ \TanGCE{} (with HPO) and the proposed closed-form pipeline (no HPO) appear in those tables to make the comparison explicit.

\paragraph{Why this is fair to baselines.}
The trust-region radius $\varepsilon$ is one global scalar shared across all panels and concepts; in contrast, the $\lambda$-HPO grid gives baseline methods a per-panel selection that strictly subsumes any single-$\varepsilon$ choice. Reporting fixed $\varepsilon=0.1$ for the proposed pipeline therefore understates rather than overstates its advantage relative to baselines that are allowed per-panel HPO.
