% Introduction
\section{Introduction}
\label{sec:intro}

Neural representations encode many attributes of their inputs: semantic content, style, demographic information, task labels, and other latent factors. In many applications, we may want to remove one such attribute, for example a protected or spurious concept (e.g., gender and toxicity), while preserving everything else the representation contains \citep{dearteaga-etal-2019-bias,avitan-etal-2025-practical}. This is the central tension in \textit{concept erasure}: removing the target information while unnecessarily damaging other control concepts, known as \textit{surgicality} \cite{singh-ravfogel-2024-representation-surgery}.

The surgicality requirement makes erasure especially difficult. A target concept is rarely encoded in isolation. It may be entangled with many other factors, most of which are unlabeled and therefore cannot be explicitly protected. Moreover, erasure must address \emph{recoverability}: even when a chosen classifier is driven to chance, a freshly trained classifier on the edited representations may still recover the protected attribute \citep{elazar-goldberg-2018-adversarial}. Linear erasure methods provide simple and effective closed-form interventions \citep{ravfogel-etal-2020-null,belrose-etal-2023-leace}, but they only remove linearly accessible information and do not generally block nonlinear probes \citep{ravfogel-etal-2022-adversarial}. Nonlinear erasure, however, introduces a different danger: if we directly follow the gradient of a nonlinear concept scorer in the ambient representation space, the edit may remove target signal by moving the representation to a region that natural inputs never occupy.

While neural representations live in $\mathbb{R}^d$, learned representations that are produced by different models do not fill this space uniformly. Instead, they concentrate near a much smaller, region that we treat as a low-dimensional \emph{manifold} embedded in the ambient space. This observation changes the geometry of erasure: an arbitrary direction in $\mathbb{R}^d$ may be a valid ambient vector, but it need not be tangent to the manifold of natural representations. Moving in such a direction pushes the representation \emph{off the manifold}, into a region the encoder does not normally produce, where downstream computations behave unpredictably and unrelated information is distorted. This is the central reason an off-manifold step damages \emph{unknown} control concepts: most of what the representation encodes is unlabeled, but every encoded concept --- labeled and unlabeled alike --- is read out by downstream layers from the same manifold. Pushing off the manifold corrupts that shared substrate; staying on it bounds the damage by construction. We call this the \emph{Manifold Representation Hypothesis} (MRH): the natural representations of a trained model concentrate on a low-dimensional manifold $\mathcal{M} \subset \mathbb{R}^d$, and an erasure step that moves each representation \emph{along} $\mathcal{M}$ is structurally surgical, while a step with off-manifold components damages the unlabeled control concepts that ride along the same manifold. Linear erasure has a closed-form, low-rank solution that is applied uniformly to every representation: the operator targets a small number of global moments (a mean direction, a covariance asymmetry) and cannot adapt per sample. Nonlinear erasure has no analogous closed form. Without one, we must estimate per-sample directions of concept presence (typically as the gradient of a learned scorer) and apply them iteratively. The estimate has no built-in constraint to remain on the manifold, so a per-sample step that locally reduces concept signal can also push the representation off-manifold. MRH makes a directly testable prediction: if the on-manifold constraint is what makes erasure surgical, then bolting it on top of \emph{any} other erasure pipeline should reduce residual concept leakage without further damaging the controls. We confirm exactly this in §\ref{sec:tangce_on_baseline}, where stacking \TanGCE{} on top of five prior erasers (LEACE, INLP, IGBP, Obliviator, AmbGCE) lowers nonlinear-probe leakage on every panel while preserving control accuracy.

We introduce the \TanGCE{} family, a manifold-aware approach to nonlinear concept erasure that operationalises this claim. The core method, \TanGCE{}, is an iterative manifold-constrained loop: for each representation, we approximate the manifold locally by its first-order tangent space at that point, estimated via SVD on nearest-neighbor secants \citep{zhang-zha-2004-ltsa}. We project the concept-scorer gradient onto this tangent space, restricting the edit to directions the manifold actually extends along. We then rescale each tangent coordinate by the corresponding local singular value, a local-$\sigma$ preconditioner that ties the step magnitude in each direction to how far the manifold extends along that axis. Two preprocessing-augmented variants compose this core with closed-form linear stages: \TanGCEp{} prepends LEACE \citep{belrose-etal-2023-leace}, exhausting the first-moment mean difference; \TanGCEpp{} further prepends a rank-2 projection that removes the leading second-moment covariance asymmetry. The combined effect is that every nonlinear edit is an on-manifold step: target signal is removed, and the unlabeled control concepts that ride along the same manifold are protected by construction.

We test the \TanGCE{} family on text and vision tasks: 13 language models spanning 0.5B--27B parameters on three target concepts (sycophancy, gender bias, safety), and CLIP image embeddings on 40 face-attribute concepts under two control regimes. We operationalize surgicality on every setting by fixing a \emph{control-damage budget}, the maximum allowed drop in held-out control-concept probe accuracy, and reporting how much target-concept signal each method removes within it (measured by a fresh nonlinear probe). \TanGCEpp{} (the full pipeline) beats the strongest published baseline at every budget we test on the NLP grid, and \TanGCEp{} (LEACE preprocessing + manifold loop) is the headline configuration on CelebA, all under one fixed set of hyperparameters with no per-setting tuning.

Our contributions are:
\begin{enumerate}
    \item We name and empirically test the \emph{Manifold Representation Hypothesis} (MRH): that on-manifold edits are surgical by construction. Beyond our own pipeline, we show that adding the manifold-constrained step on top of five unrelated prior erasers (LEACE, INLP, IGBP, Obliviator, AmbGCE) reduces residual leakage while preserving control accuracy (§\ref{sec:tangce_on_baseline}) --- direct evidence that the manifold constraint, not any one preprocessor, is doing the work.
    \item We introduce the \TanGCE{} family: a manifold-aware iterative erasure step (\TanGCE{}), and two preprocessing-augmented variants (\TanGCEp{} = LEACE + \TanGCE{}; \TanGCEpp{} = LEACE + rank-2 CovMatch + \TanGCE{}) that ablate the contribution of each closed-form preprocessing stage.
    \item We propose a local-$\sigma$ preconditioner that scales tangent-coordinate steps by the local singular spectrum, making the trust region sensitive to the manifold's anisotropic geometry.
\end{enumerate}
