% Results
\section{Results}
\label{sec:results}

We evaluate eight erasure methods plus a Majority Class reference baseline across three concept domains---sycophancy, gender (Bias in Bios), and safety---at layer~8 of Qwen2.5-0.5B. Each method's erasure strength $\lambda^*$ is selected from the same shared validation protocol described in Section~\ref{sec:evaluation}. Our main metrics are erasure precision, measured by retrained probe accuracy on the target concept, and surgicality, measured by change in control-task performance. The main comparison is between ambient-space editing and tangent-constrained editing.

\input{tables/full_comparison}

\subsection{\TanGCE{} Achieves the Best Erasure Precision Across All Domains}

Table~\ref{tab:full_comparison} shows that \TanGCE{} achieves the lowest nonlinear leakage across all three domains: 0.602 on sycophancy (vs.\ 0.681 for LEACE, 0.703 for \AmbGCE{}, 0.748 for IGBP), 0.618 on gender (vs.\ 0.675 for IGBP, 0.701 for \AmbGCE{}), and 0.663 on safety (vs.\ 0.759 for IGBP, 0.778 for \AmbGCE{}). The Majority Class row gives the trivial label-imbalance floor, showing what it would mean to erase by collapsing to the dominant class rather than preserving useful structure. Linear methods (Linear Projection, INLP) and the closest prior nonlinear baseline (IGBP) remain well above \TanGCE{}, confirming that concept information is predominantly encoded nonlinearly and that ambient-space updates alone are not enough.

\subsection{\TanGCE{} Remains Surgical on Control Concepts}

Table~\ref{tab:full_comparison} reports $\Delta$Ctrl, the change in downstream control task accuracy relative to clean representations. The key point is not just that some baselines can lower leakage, but how they do it. The random baselines and IGBP can reduce recoverability while still inducing substantially larger collateral damage than \TanGCE{} in the harder settings, indicating that they are more willing to distort the representation off its natural support. For gender (profession classification), \TanGCE{} drops only $-0.8$pp (0.531 vs.\ 0.539 clean), whereas IGBP drops $-1.5$pp and the random per-sample baseline is worse still. On safety, \TanGCE{} \emph{improves} helpfulness ($+3.9$pp), while IGBP also lowers leakage but with a larger representation shift and a different utility profile. The Majority Class row is a reference floor rather than an intervention, so the control metric is not applicable.

\subsection{The Lambda Sweep: On-Manifold Updates Support Larger Interventions}
\label{sec:lambda_sweep}
\input{tables/lambda_sweep}

Table~\ref{tab:lambda_sweep} isolates the effect of step size by running only \AmbGCE{} and \TanGCE{} at three fixed $\lambda$ values ($0.5, 4.0, 8.0$) on the same three domains as above (Qwen2.5-0.5B, layer~8, same train/val/test splits and fresh-probe evaluation protocol as in Section~\ref{sec:evaluation}). As $\lambda$ increases, \AmbGCE{} becomes unstable: its nonlinear leakage worsens sharply at high step sizes, rising to 0.734 on sycophancy, 0.702 on gender, and 0.770 on safety at $\lambda{=}8$. \TanGCE{} behaves in the opposite way, improving as $\lambda$ increases and reaching its best precision at the largest tested step size in all three domains.

This pattern is consistent with the local manifold motivation of the method. If the ambient update includes components that are weakly supported by the local representation geometry, increasing $\lambda$ can make the edit less effective under the nonlinear recoverability metric. Projecting into the estimated tangent space removes those components at first order, which can make larger steps more useful in practice. We treat this as supportive evidence for the manifold-support view, not as a complete characterization of the leakage-utility tradeoff.


\subsection{Cross-Modal Generalization: CelebA Visual Attributes}
\label{sec:celeba}

To test whether \TanGCE{} generalizes beyond language model representations, we apply all seven erasure methods to pooled CLIP ViT-B/32 image embeddings (768-dim) of CelebA face images (see Section~\ref{sec:evaluation} for preprocessing and split details). CelebA provides 40 binary facial attributes, giving a rich surgicality profile: when erasing one attribute, the remaining attributes serve as controls. The same validation-HPO protocol ($\lambda^* \in \{0.5, 1.0, 2.0, 4.0, 8.0\}$ by lowest validation NL probe accuracy, hidden 256, 1500 steps) is applied uniformly across all methods.

We treat this cross-modal setting as a generality check rather than the primary evidence for the paper's central claim.

\input{tables/celeba_clip_fair}

Table~\ref{tab:celeba_clip_fair} shows results on three representative concepts (Male, Smiling, Blond\_Hair), each with 5 low-correlation control attributes. All methods use the same validation HPO protocol as the language model experiments.

\input{tables/celeba_clip_all40_summary}

Table~\ref{tab:celeba_clip_all40} summarizes results averaged over all 40 CelebA attributes. The median selected $\lambda^*$ column confirms that \TanGCE{} consistently selects higher $\lambda$ values than other methods, enabled by the tangent projection's attenuation of off-manifold displacement, while maintaining competitive control accuracy. Per-concept breakdowns for all 40 attributes (nonlinear leakage and mean $\Delta$Ctrl per method) are reported in Appendix~\ref{sec:appendix_celeba_per_concept}, Tables~\ref{tab:celeba_clip_all40_per_concept_leak} and~\ref{tab:celeba_clip_all40_per_concept_ctrl}.

\subsection{Multi-Model Generalization}
\label{sec:multi_model}

To assess whether the ambient-vs-tangent comparison is specific to one backbone, we evaluate across five model families: Qwen2.5 (0.5B, 1.5B, 3B), Gemma-2 (2B, 9B, 27B), Llama-3.2 (1B, 3B), Mistral-7B, and Gemma-3 (1B, 4B, 12B, 27B). Each model is evaluated at the layers corresponding to 25\%, 50\%, and 75\% of its depth; we report the mean across these three layers (standard deviations in Appendix Tables~\ref{tab:appendix_all_methods_sycophancy}--\ref{tab:appendix_all_methods_safety}). We compare only \TanGCE{} and \AmbGCE{}, the two methods that differ solely in their gradient constraint, isolating the effect of tangent-space projection. The validation HPO protocol uses $\lambda^* \in \{0.25, 0.5, 1.0, 2.0, 4.0, 8.0, 16.0, 32.0\}$ selected by lowest validation nonlinear probe accuracy. Per-layer results are in Table~\ref{tab:multi_model_per_layer}; the main table reports the corresponding mean results for all completed models.

\textbf{Bold} marks the best cell per column (lowest NL or $|\Delta$Ctrl| closest to zero) across each (model, method) row.

\input{tables/multi_model_comparison}

\TanGCE{} achieves lower nonlinear leakage than \AmbGCE{} across all completed models and all three concept domains, including Gemma-3-27B. The advantage is consistent across concept types: on gender (Bias in Bios), the gap is particularly pronounced---\TanGCE{} reduces NL by $0.196$--$0.303$ points relative to \AmbGCE{} across families (e.g., Mistral-7B: $0.912 \to 0.634$; Llama-3.2-3B: $0.926 \to 0.646$; Gemma-2-2B: $0.810 \to 0.573$; Gemma-3-27B: $0.918 \to 0.659$). On safety, gaps range from $0.108$ to $0.168$ points across the originally reported families, and reach $0.145$ on Gemma-3-1B, $0.156$ on Gemma-3-4B, and $0.165$ on Gemma-3-27B. On sycophancy, gaps are smaller but still present, including Gemma-3-27B ($0.581 \to 0.580$). Just as importantly, the baselines that appear strongest on raw nonlinear leakage often do so by distorting the representation itself: the random baselines can lower NL, but their control-task damage is typically about an order of magnitude larger than \TanGCE{} on the same model family. IGBP is the strongest prior nonlinear baseline in this comparison family, but it remains consistently weaker than \TanGCE{} on nonlinear leakage while still incurring larger control degradation in the harder settings. $\Delta$Ctrl remains near zero for \TanGCE{} throughout most families; Gemma-3-27B remains a caveat because its gender edit takes a larger control hit ($-0.129$), even though the removal advantage remains large.

\paragraph{Interpretation.}
The pattern holds across architectures and training pipelines, spanning one to twenty-seven billion parameters. Within the Qwen2.5 family, AmbGCE NL increases monotonically with scale ($0.806 \to 0.812 \to 0.861$ on sycophancy), while \TanGCE{} remains consistently lower. A similar trend holds within Gemma-2 ($0.656 \to 0.746 \to 0.759$) and Llama-3.2 ($0.831 \to 0.882$). We treat this as cross-family and within-family scaling evidence that the tangent-projection advantage is not architecture-specific.
